
Each time the rational programmer debugs a program, it is validation for
programming languages researchers; their blame assignment mechanisms
provide information that points towards the source of the problem.
Moreover, their slogans and theorems help programmers connect the dots.
In contrast, when the rational programmer fails, it speaks for the limited
predictive power of programming languages theory and the misleading
projections of language design.

In other words the rational programmer implies a methodology for
evaluating the design of programming languages features such as blame
assignment mechanisms.  In general terms, the methodology consists of four
pieces:

\begin{itemize} 

 
  \item The description of a  rational programmer as an algorithmic
    procedure that examines a program and performs an action.  Typically
    the action of the rational program is consistent with the theory of
    the features under examination.

 
  \item A hypothesis about whether a sequence of actions of the rational
    programmer leads to a desired outcome.

  \item A large-scale experiment that tests the hypothesis on
    representative programs. 


  \item The evaluation of the features based on the outcome of the
    experiment.    

\end{itemize}

In specific terms for blame assignment mechanisms, the methodology
imagines:  

\begin{itemize}
 
  \item The rational programmer as a procedure that consumes a program and
    the error message from evaluation of the program and uses the error
    message to select the next component to add types.

  \item The rational programmer hypothesis as  research question whether
    for all programs with type-level mistakes a series of steps of the
    rational programmer always leads to a type error that points to the
    component that contains the mistake. 

  \item The experiment as the attempts of the rational programmer to debug
    a large number of programs with type-level mistakes. 


  \item The evaluation of a blame assignment mechanism based on whether
    the information it provides is critical for the outcome of the
    experiment.  

\end{itemize}




