

%% -----------------------------------------------------------------------------
\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}

The Natural semantics assigns blame to exactly one boundary.  A blame assignment
has the following specific meaning: the typed component may make incorrect type
assumptions about the untyped component in its interface, or the correct
interface exposes a bug in the untyped component. Our setup rules out the first
alternative (but see section~\ref{sec:conclusion}), and therefore the rational
programmer extends the trail to a scenario that swaps out the untyped component
for its typed counterpart.

We can turn this description into a formal definition of the trail. 
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
the rational programmer gets no hints from the gradual type system on how to
continue the search for the bug.

While a successful Natural blame trail indicates that it 
pays off to heed blame assignments during debugging the trail's root, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run-time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the Natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
\end{quote}

Same as for success for a Natural blame trail, we define success for a
Natural exception trail.
Putting the definitions for the two modes together, we can now compare the usefulness of blame 
with that for mere exceptions for debugging a scenario in the context of Natural semantics.
\begin{quote}
\it 
  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
  Natural blame is \emph{more useful} than Natural exceptions for
  debugging $\conf_0$ iff 
  the Natural blame trail 
  that starts at $\conf_0$ is successful while the Natural exception trail that
  starts at $\conf_0$ is failing.
\end{quote}
